Seminar: Programming models and Code Generation\\ Winter Term 2014/2015\\ {\bf Quasi deterministic parallel programming using LVars}

### 2015-01-07 draft.tex
new

The purpose of a deterministic-by-construction programming model is to rule out programs like the following from being written, or force them to behave deterministically. \cite{lindsey}Written in a hypothetical parallel language. 
\begin{verbatim}
let _ = put lv 3 in  /write 3 in lv/
let par v = get lv  /read from lv/
        _ = put lv 4  /write 4 in lv/
	      in v   /final value is v/
\end{verbatim} 
The \textit{let par} expression which has \textit{fork-join semantics}, evaluates the two sub-expressions in parallel. The value of $v$ is the value of the entire program. Depending on the scheduler \textit{get} $lv$ or \textit{put} $lv\ 4$ can run first, $v$ might be either $3$ or $4$ during multiple runs giving rise to non determinism.\par
A popular approach for enforcing determinism is to require that variables can only be assigned to once, giving us a single-assignment language. Our program would be invalid because it tries to write to $lv$ twice. Single-assignment variables with blocking read semantics, often known as IVars, turn up in all kinds of deterministic parallel systems. But the single-assignment rule disallows a number of programs that we would like to permit such as writing the same value to an IVar twice. As another example, consider the problem of filling in various parts of a data structure in parallel. \cite{lindsey}Say that $lv$ contains a pair, and we are writing to $lv$ twice, with each write filling in one component of the pair. Please note that $x=\bot$ below.\par 
\begin{verbatim}
let par  _ = put lv (4,x)
         _ = put lv (x,3) /error/
         in get lv
\end{verbatim}  
Read from $lv$ is possible once both writes have completed, so it will always be (4, 3). But single-assignment rules out this program(although we could fake it by using a distinct IVar for each component of the pair, but with more sophisticated data structures it becomes difficult to fake a solution with IVars.)\par

@@

\section{LVars and LVish}
\subsection{LVars}
An LVar is a memory location that can be shared between multiple threads whose contents can only grow bigger over time, for some definition of “bigger” that the user of the LVar gets to specify. \par
LVars, short for lattice variables, are a generalization of IVars, which are variables that can only be written to once. (The “I” is for “immutable”.) LVars, on the other hand, allow multiple writes, as long as those writes are monotonically increasing with respect to a user-specified lattice. Meanwhile, LVar reads have to specify a threshold set of possible states. A read will block until one of the states in the threshold set has been reached, and then return that state. Information can never be removed and the order in which it is added is not observable. IVars turn out to be a special case of LVars. 
The property of monotonically increasing write operations is enforced by the language. The semantics of $put$ says that a write to a location is the least upper bound of the new value and the current value. 
\begin{verbatim}
let par _ = put num 3
        _ = put num 2
        x = get num
  in x  /*could be 2 or 3*/
\end{verbatim}
Monotonic writes alone aren't enough to ensure determinism. The above program is non-deterministic where num's notion of “bigger” is $\geq$. Despite writes always being monotonically increasing. $Get$ might read either $2$ or $3$, depending on the order of $put$ or $get$ operations determined by the scheduler. With an LVar model this behavior changes.\par
\begin{itemize}
\item To maintain determinism, the LVars model restricts the $get$ operation. Instead of reading the LVar’s exact value, it reads one of a set\{threshold set\} of lower bounds on its value that are given as an extra argument to $get$ not shown in the example. The $get$ operation will block until the LVar reaches a value that is at or above one of those lower bounds. For the above example, the threshold set will just be \{$3$\}. $Get$ operation is called a threshold read, because the value it returns is the threshold we have crossed.
\item Once num value is atleast $3$, it will remain at or above $3$ forever. As long as we only access them through the $put$ or $get$ interface, LVars are thread-safe: we can safely share them between threads without introducing unintentional non-determinism. That is, $puts$ and $gets$ can happen in any order, without changing the value that a program evaluates to.
\end{itemize}
%\begin{center}
%\textit{Monotonically increasing writes + threshold reads = deterministic parallelism}
%\end{center}
The resulting model is general enough to subsume the existing single-assignment model. The user-specified set just needs to be a \textit{join-semilattice}, which is a partially ordered set in which every two elements have a least upper bound.

@@

Freezing does not commute with writes. If a freeze is interleaved before such a write, the write will raise an exception; if it is interleaved afterwards, the program will proceed normally. Although it appears like determinism is lost, these programs still satisfy the property of Quasi determinism that is they will always output the same result or an error.

@@

The quiesce operation blocks until all of the callbacks running in a particular handler pool, in this case $h$, are done running. 
where $lv$ is an LVar, $Q$ is an event set, and $f$ is a callback. Handler pools are created with the $newPool$ function, and handlers are registered with $addInPool$. Quiescence of a handler is a non-monotonic property: it can move in and out of quiescence as more puts to an LVar occur. There is no risk to quasi-determinism, however, because quiesce does not yield any information about which events have been handled, any such questions must be asked through Lvar $get$.

The freeze operation is a non-blocking read that lets us find out the exact contents of an LVar. After freezing, any further writes that would change LVar's state will raise an exception! A safe practice is to quiece before freezing For a program that performs freeze, Only two possible outcomes are guaranteed: the deterministic result of all the effects, or a write-after-freeze exception that can assist in debugging the synchronization bug. This property is called quasi-determinism, and the Lvish model is
quasi-deterministic.

Freezing introduces quasi-determinism to the programming model, because we might freeze before we have quiesced, or because our thread that's freezing and quiescing might be racing with some other thread to write into a shared LVar. But, if it is ensured that freezes only ever happen after all writes have completed, then the computations will be deterministic, because there wouldn't be any write-after-freeze races. This is enforced at the implementation level which takes care of quiescing and
freezing at the end of a computation. In the LVish library, this is provided by the runParThenFreeze function.\\
\cite{lkuper2}contains a quasi-deterministic, parallel, call-by-value $\lambdaLVish$-calculus extended with a store containing LVars. It extends the $\lambdaLVar$ calculus to support event handlers and freezing. Rather than modeling the full ensemble of event handlers, handler pools, quiescence, and freezing as separate primitives, it instead formalizes the \emph{“freeze-after”} pattern which is a simpler primitive combining all others.

>>>>

The purpose of a deterministic-by-construction programming model is to rule out programs like the following from being written, or force them to behave deterministically. Written in a hypothetical parallel language~\cite{lindsey}. 
\begin{verbatim}
let _ = put lv 3 in  /write 3 in lv/
let par v = get lv  /read from lv/
        _ = put lv 4  /write 4 in lv/
	      in v   /final value is v/
\end{verbatim} 
The \textit{let par} expression which has \textit{fork-join semantics}, evaluates the two sub-expressions in parallel. The value of $v$ is the value of the entire program. Depending on the scheduler \textit{get} $lv$ or \textit{put} $lv\ 4$ can run first, $v$ might be either $3$ or $4$ during multiple runs giving rise to non determinism.\par
A popular approach for enforcing determinism is to require that variables can only be assigned to once, giving us a single-assignment language. Our program would be invalid because it tries to write to $lv$ twice. Single-assignment variables with blocking read semantics, often known as IVars, turn up in all kinds of deterministic parallel systems. But the single-assignment rule disallows a number of programs that we would like to permit such as writing the same value to an IVar twice. As another example, consider the problem of filling in various parts of a data structure in parallel. \cite{lindsey}Say that $lv$ contains a pair, and we are writing to $lv$ twice, with each write filling in one component of the pair. Please note that $x=\bot$ below.\par 
\begin{verbatim}
let par  _ = put lv (4,x)
         _ = put lv (x,3) /error/
         in get lv
\end{verbatim}  
Read from $lv$ is possible once both writes have completed, so it will always be (4, 3). But single-assignment rules out this program(although we could fake it by using a distinct IVar for each component of the pair, but with more sophisticated data structures it becomes difficult to fake a solution with IVars.)\par

@@

\section{LVars and LVish}
\subsection{LVars}
An LVar is a memory location that can be shared between multiple threads whose contents can only grow bigger over time, for some definition of “bigger” that the user of the LVar gets to specify. \par
LVars, short for lattice variables, are a generalization of IVars, which are variables that can only be written to once. (The “I” is for “immutable”.) LVars, on the other hand, allow multiple writes, as long as those writes are monotonically increasing with respect to a user-specified lattice. Meanwhile, LVar reads have to specify a threshold set of possible states. A read will block until one of the states in the threshold set has been reached, and then return that state. Information can never be removed and the order in which it is added is not observable. IVars turn out to be a special case of LVars. 
The property of monotonically increasing write operations is enforced by the language. The semantics of $put$ says that a write to a location is the least upper bound of the new value and the current value. 
\begin{verbatim}
let par _ = put num 3
        _ = put num 2
        x = get num
  in x  /*could be 2 or 3*/
\end{verbatim}
Monotonic writes alone aren't enough to ensure determinism. The above program is non-deterministic where num's notion of “bigger” is $\geq$. Despite writes always being monotonically increasing. $Get$ might read either $2$ or $3$, depending on the order of $put$ or $get$ operations determined by the scheduler. With an LVar model this behavior changes.\par
\begin{itemize}
\item To maintain determinism, the LVars model restricts the $get$ operation. Instead of reading the LVar’s exact value, it reads one of a set\{threshold set\} of lower bounds on its value that are given as an extra argument to $get$ not shown in the example. The $get$ operation will block until the LVar reaches a value that is at or above one of those lower bounds. For the above example, the threshold set will just be \{$3$\}. $get$ operation is called a threshold read, because the value it returns is the threshold we have crossed.
\item Once num value is atleast $3$, it will remain at or above $3$ forever. As long as we only access them through the $put$ or $get$ interface, LVars are thread-safe: we can safely share them between threads without introducing unintentional non-determinism. That is, $puts$ and $gets$ can happen in any order, without changing the value that a program evaluates to.
\end{itemize}
%\begin{center}
%\textit{Monotonically increasing writes + threshold reads = deterministic parallelism}
%\end{center}
The resulting model is general enough to subsume the existing single-assignment model. The user-specified set just needs to be a \textit{join-semilattice}, which is a partially ordered set in which every two elements have a least upper bound.

@@

Freezing does not commute with writes. If a freeze is interleaved before such a write, the write will raise an exception; if it is interleaved afterwards, the program will proceed normally. Although it appears like determinism is lost, these programs still satisfy the property of quasi determinism that is they will always output the same result or an error.

@@

The $quiesce$ operation blocks until all of the callbacks running in a particular handler pool, in this case $h$, are done running, $lv$ is an LVar, $Q$ is an event set, and $f$ is a callback. Handler pools are created with the $newPool$ function, and handlers are registered with $addInPool$. Quiescence of a handler is a non-monotonic property: it can move in and out of quiescence as more puts to an LVar occur. There is no risk to quasi-determinism, however, because quiesce does not yield any information about which events have been handled, any such questions must be asked through $get$ operation.

The freeze operation is a non-blocking read that lets us find out the exact contents of an LVar. After freezing, any further writes that would change LVar's state will raise an exception! A safe practice is to $quiesce$ before freezing For a program that performs freeze, Only two possible outcomes are guaranteed: the deterministic result of all the effects, or a write-after-freeze exception that can assist in debugging the synchronization bug. This property is called quasi-determinism, and the Lvish model is
quasi-deterministic.

Freezing introduces quasi-determinism to the programming model, because we might freeze before we have quiesced, or because our thread that's freezing and quiescing might be racing with some other thread to write into a shared LVar. But, if it is ensured that freezes only ever happen after all writes have completed, then the computations will be deterministic, because there wouldn't be any write-after-freeze races. This is enforced at the implementation level which takes care of quiescing and
freezing at the end of a computation. In the LVish library, this is provided by the $runParThenFreeze$ function.\\
\cite{lkuper2}contains a quasi-deterministic, parallel, call-by-value $\lambdaLVish$-calculus extended with a $store$ containing LVars. It extends the $\lambdaLVar$ calculus to support event handlers and freezing. Rather than modeling the full ensemble of event handlers, handler pools, quiescence, and freezing as separate primitives, it instead formalizes the \emph{“freeze-after”} pattern which is a simpler primitive combining all others.
